The impact of gas bulk rotation on the Lyman-$\alpha$ line

### 2013-11-01 paper/ms.tex
new edits in the discussion and conclusions

\begin{figure}
  \includegraphics[scale=0.35]{OF_spheres.png}
  \caption{The two solid circles show the two different spherical
    off-centred emitting  regions. The angular velocity vector is
    defined to be along the $\hat{\i}$ direction and the observer is
    along the $\hat{\j}$ direction (where the reader is). 
    \label{fig:OCspheres}} 
\end{figure}

The model with homogeneous radiation sources allows to study the
effect of radiation sources distributed with a different symmetry than
the gas. We select photons emitted in off-centred spheres 
with respect to the gas distribution.

Figure \ref{fig:OCspheres} shows the two spherical regions we
define. Each has a radius of $0.5$R and can have to possible centres
positions $C$ along the $\hat{\i}$ direction: $\pm C\hat{\i}$. We also
place an observer along the $\hat{\j}$ direction, such that the
emitting sphere centred on $+C\hat{\i}$ moves away from the observer
and the sphere centred on $+C\hat{\i}$ moves towards the observer. We
consider two values for $C=(0.25,0.50)R$.

We study first the angular deviations of the outgoing photons with respect
to an isotropic flux. We follow \cite{Zheng2013} to estimate the flux
as a function of the angle $\Theta$ formed by the outgoing photons and
a fixed direction (the $\vec{k}$ direction in this case). This is
expressed by

@@

Figure \ref{fig:OC_profiles} show the spectra for the two off-centered
spheres located at $C=\pm0.5R$.  The most interesting effect is that
emission is that in  some models ($\tau = 10^6, V=100,200$\kms and $\tau =
10^7, V=200,300$\kms) the line presents three peaks (blue-shifted,
centred, red-shifted). However, we do not find any asymmetry in the line 
produced by the off-center emission. As in the homogeneous / central
model the line gets broader as the rotational velocity increases.

@@

Gas bulk rotation has a noticeable effect on the morphology of the
 \ly line. In this section we discuss the implications of these
 findings for the interpretation of available observational data. Our
 discussion will focus on the qualitative aspects of diverse
 morphological features observed by \cite{Kulas12} ans \cite{Yamada2012}

We follow a classification of $Ly\alpha$ profiles made by
\cite{Kulas12}. They presented observational results for star forming
galaxies at $z=2$-$3$ that showed multiple peaks in their emission.
They classified the observed lines into five groups. In Group I the
red peak is stronger than the blue, while in Group II the blue peak is
stronger, in both cases the peaks are symmetrically located around the
line center. In Group III the blue peak also dominates but there is an
overall redward shift of the zero point. Group IV presents two similar
peaks symmetrically located around the zero point. Group V presents
three peaks.

The asymmetries in Groups I, II  and III cannot be fully explained by
rotation. However, some galaxies in those groups have a high flux between
the two peaks; an effect that can be induced by rotation. The same
argument applies to galaxies in the group IV in which the flux
in between of the two peaks is high ($ \geq 70 \%$ of the intensity in
the highest peak). Finally, the triple peaked lines in Galaxies in
Group V can be easily reproduced by rotation with off-centred
emission.

Another relevant sample of lines was presented by \cite{Yamada2012} 
for 91 LAEs at $z=3.1$. They find that $\sim 33 \%$ of the galaxies present 
a single peak profile. Some of the peaks have a strong symmetry, while
others do not. The symmetric single peaked profile can be explained by
rotation effects, while the asymmetric cases naturally arise in
outflow/inflow models \citep{Verhamme2008, Dijkstra06}. Among the 91
LAEs, 8 reveal triple peaked profiles with a central peak and two
smaller peaks at the redder and bluer side, a feature that is present
in our off-centre emission models with rotation.

The presence of single peaked profiles has been associated to
inflow/outflow dynamics \citep{XXX}. In this paper we show that gas
bulk rotation can also be considered as a probable origin for that
behaviour, provided that the observed single peak is highly
symmetric. Similarly, in the case of double peaked lines with a high
level of flux at the line center, rotation also deserves to be considered in
the pool of possible bulk flows responsible for that feature,
specially if the two peaks have similar intensities. The case of
triple peaked lines is another clear feature of gas rotation, this
time with the additional condition of an off-centred emission.

@@

In this paper, we have estimated the effects of gas bulk rotation on
the morphology of the Lyman $\alpha$ line in star forming
galaxies. The results are based on the study of a homogeneous sphere
rotating as a solid body. We explored a range of models by varying
the rotation speed, hydrogen optical depth, dust optical depth and
initial distribution of Ly$\alpha$ photons with respect to the gas
density. As a cross-validation, we obtain our results from two
independently developed Monte-Carlo radiative transfer codes.

Our main result is that rotation clearly impact the \ly line
morphology. Double peaked lines can make transitions into single
peaked lines as the rotational velocity increases, off-centred
can result into triple peaked lines.

@@

As many observational spectra Ly$\alpha$ emission line (Kulas et al,
Yamada et al) is double  peaked, these peaks provide important information
concerning gas kinematics and geometry,  which can be partially
explained with inflows/outflows of gas content.  We study the effect
of rotation in the position of this peaks, and we find  that the
position of the maxima does change with rotation for the homogeneous
models when the double peak merged into a single peak as velocity
increase. This effect is not seen for the central distribution when
the double peak remains constant as the velocity increase. We also
find that there is no dependency in the observer viewing angle with
the maxima position.

>>>>

\begin{figure}
  \includegraphics[scale=0.35]{OF_spheres.png}
  \caption{The two solid circles show the two different spherical
    off-centred emitting  regions. The angular velocity vector is
    defined to be along the $\hat{\i}$ direction and the observer is
    along the $\hat{\j}$ direction (where the reader is) (THE ANGLE
    THETA IS WRONG. IT IS THE ANGLE WITH THE OBSERVER, RIGHT?). 
    \label{fig:OCspheres}} 
\end{figure}

The raw data from the model with homogeneous radiation sources allows
to study the effect of radiation sources distributed with a different
symmetry than the gas. We can select photons emitted in off-centred spheres 
with respect to the gas distribution and check whether this produces
asymetries in the line.

Figure \ref{fig:OCspheres} shows the two spherical regions we
define to test for this effect. Each has a radius of $0.5$R and can
have to possible centres positions $C$ along the $\hat{\i}$ direction:
$\pm C\hat{\i}$. We also place an observer along the $\hat{\j}$
direction, such that the emitting sphere centred on $+C\hat{\i}$ moves
away from the observer and the sphere centred on $+C\hat{\i}$ moves
towards the observer. We consider two values for $C=(0.25,0.50)R$.

We study first the angular deviations of the outgoing photons with respect
to an isotropic flux. We follow \cite{Zheng2013} to estimate the flux
as a function of the angle $\Theta$ formed by the outgoing photons and
a the observer. This is expressed by

@@

Figure \ref{fig:OC_profiles} show the spectra for the two off-centered
spheres located at $C=\pm0.5R$.  In this case we only use photons that
have outgoing directions aligned with the observer. The most
interesting effect is that emission is that in  some models ($\tau =
10^6, V=100,200$\kms and $\tau = 10^7, V=200,300$\kms) the line
presents three peaks (blue-shifted,centred, red-shifted). However, we
do not find any asymmetry in the line produced by the off-center
emission. As in the homogeneous / central model the line gets broader
as the rotational velocity increases.

@@

Gas bulk rotation has a noticeable effect on the morphology of the
 \ly line. In this section we discuss the implications of these
 findings for the interpretation of available observational data. Our
 discussion focuses on the qualitative aspects of diverse
 morphological features observed by \cite{Kulas12} ans \cite{Yamada2012}

We follow a classification of \ly profiles made by
\cite{Kulas12}. They presented observational results for star forming
galaxies at $z=2$-$3$ with multiple peaks in their \ly emission.
They classified the observed lines into five groups. In Group I the
red peak is stronger than the blue, while in Group II the blue peak is
stronger, in both cases the peaks are symmetrically located around the
line center. In Group III the blue peak also dominates but there is an
overall redward shift of the zero point. Group IV presents two similar
peaks symmetrically located around the zero point and Group V galaxies show
three peaks.

The asymmetries in Groups I, II  and III cannot be explained by
rotation. However, some galaxies in those groups have a high flux between
the two peaks; an effect can be induced by rotation. The same
argument applies to galaxies in the group IV in which the flux
in between of the two peaks is high ($ \geq 70 \%$ of the intensity in
the highest peak). Finally, the triple peaked lines in Galaxies in
Group V can be easily reproduced by rotation with off-centred
emission given the right amount of asymmetry between the emitting
region and neutral hydrogen distribution.

Another relevant sample of observations was presented by \cite{Yamada2012} 
for 91 LAEs at $z=3.1$. They find that $\sim 33 \%$ of the galaxies present 
a single peak profile. Some of the peaks have a strong symmetry, while
others do not. The symmetric single peaked profile can be explained by
rotation effects, while the asymmetric cases naturally arise in
outflow/inflow models \citep{Verhamme2008, Dijkstra06}. Among the 91
LAEs, 8 reveal triple peaked profiles with a central peak and two
smaller peaks at the redder and bluer side, a feature that is present
in our off-centre emission models with rotation.

The presence of single peaked profiles has been associated to
inflow/outflow dynamics \citep{XXX}. In this paper we show that gas
bulk rotation can also be considered as a probable origin for that
behaviour, provided that the observed single peak is highly
symmetric. Similarly, in the case of double peaked lines with a high
level of flux at the line center, rotation also deserves to be considered in
the pool of possible bulk flows responsible for that feature,
specially if the two peaks have similar intensities. The case of
triple peaked lines is another clear feature of gas rotation under the
additional condition of an off-centred emission.

@@

In this paper, we have estimated the effects of gas bulk rotation on
the morphology of the Lyman $\alpha$ line in star forming
galaxies. The results are based on the study of an homogeneous sphere
with solid body rotation. We explored a range of models by varying
the rotational speed, hydrogen optical depth, dust optical depth and
initial distribution of Ly$\alpha$ photons with respect to the gas
density. As a cross-validation, we obtain our results from two
independently developed Monte-Carlo radiative transfer codes.

Our main result is that rotation clearly impact the \ly line
morphology. Double peaked lines can make transitions into single
peaked lines as the rotational velocity increases. Additional
off-centred emission can result into triple peaked lines.

@@

We also studied the effect of rotation in the position of the peaks in
the line. We find that in the regions with central emission the peaks
do not shift as rotation increases. However, in the homogeneous models
the double peak merges into a single peak.
